---
title: "Rainfall and roof shape"
output:
  github_document: default
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
#run the outcommented render command below to generate both .html and .md output, the latter of which will be rendered nicely on GitHub:
#rmarkdown::render("analysis.Rmd", output_format = "all",encoding="UTF-8")
knitr::opts_chunk$set(fig.path='figs/', echo=TRUE, warning=FALSE, message=FALSE)
```

## Intro
This is a cross-cultural test of a hunch about the relation of roof shape to rainfall (briefly, lots of rain negatively selects against flat roofs). If correct, this is an example of culture-environment interactions. A possible counterexample in Ghana is of special interest, as it provides an opportunity to explore the relation between generalizations derived from large-scale cross-cultural datasets like D-PLACE and accounts grounded more directly in ethnographic fieldwork. 

# Preliminaries
Clear workspace, check for /in/ and /out/ directories (create them if needed), and load packages (install them if needed)

```{r preliminaries, results="hide"}

list.of.packages <- c("tidyr","ggthemes","lme4","extremevalues","mapproj","sp","rgeos","knitr","kableExtra")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)>0) install.packages(new.packages)
lapply(list.of.packages, require, character.only=T)
rm(list.of.packages,new.packages)

# Botero 2014
source('in/MuMAC.09Sep14.R')
source('in/MuMAC.09Sep14_distance.R')

```

# Hypothesis

*Main hypothesis*: Roof type is related to precipitation such that round/sloping roofs more likely when there is much rain. Note the directionality: we expect few flat roofs in high rainfall societies, but we have no predictions for what happens in low rainfall societies. In causal terms, high rainfall selects *against* flat roofs, but low rainfall does not select *for* them.

# Data

Load and pre-process D-PLACE data. Here's a [direct link](https://d-place.org/societies?c=%5B%7B%22variable%22:83,%22id%22:654%7D,%7B%22variable%22:83,%22id%22:655%7D,%7B%22variable%22:83,%22id%22:656%7D,%7B%22variable%22:83,%22id%22:657%7D,%7B%22variable%22:83,%22id%22:658%7D,%7B%22variable%22:83,%22id%22:659%7D,%7B%22variable%22:83,%22id%22:660%7D,%7B%22variable%22:83,%22id%22:661%7D,%7B%22variable%22:83,%22id%22:662%7D%5D&p=%5B1,2,3,4,5,6,7,8,9,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45,46,47,48,49,50,51%5D&e=%5B%5B3,%22inrange%22,%5B%220.9970%22,%22547.9482%22%5D%5D%5D) to the query on D-PLACE. Attribution:

> Research that uses data from D-PLACE should cite both the original source(s) of  the data and the paper by Kirby et al. in which D-PLACE was first presented  (e.g., research using cultural data from the Binford Hunter-Gatherer dataset: ""Binford (2001);  Binford and Johnson (2006); Kirby et al. 2016)."
> The reference list should include the date data were accessed and URL for D-PLACE (https://d-place.org), in addition to the full references for Binford (2001),  Binford and Johnson (2006), and Kirby et al. (2016)."

```{r data}
# Load data (skipping line 1, which is a note about attribution)

d <- read.csv(file="in/dplace-export.csv",skip=1, stringsAsFactors = FALSE) %>%
  # rename unwieldy var names
  dplyr::rename("roofdesc" = "Description..EA082.House.construction..shape.of.roof",
                "roofcode" = "Code..EA082.House.construction..shape.of.roof",
                "rainfall" = "Variable..Monthly.Mean.Precipitation..mm.",
                "longitude" = "Revised.longitude",
                "latitude" = "Revised.latitude",
                "references" = "References..EA082.House.construction..shape.of.roof",
                "region" = "Region.name") %>%
  # create summary variable for flat vs non-flat roofs
  mutate(flatroof = as.factor(ifelse(d$roofcode == 7,1,0)))

# as a sanity check, have a look at the roofcodes to see we're right in keeping 7 vs rest
# looks like we have only 95 societies with flat roofs and 1035 with non-flat
d %>%
  select(roofcode,roofdesc) %>%
  group_by(roofcode) %>%
  count(roofcode,roofdesc)
```
(There is also a bit of data for roof shape of secondary buildings under code EA087, but it only covers 30% of the dataset and the biases hypothesised here are more important for main dwellings so we won't use it here.)

# First exploration
So now: how does roof flatness relate to precipitation? Comparing means, we see that the average monthly rainfall for flat roof societies is 48.1 (582 annually), while for non-flat roof societies it is twice as high at 105.8 (1270 annually). To put this in context, average annual rainfall over land on earth is 59.5 monthly (715 annually) — so the mean for flat roofs seems lower and for non-flat roofs quite a bit higher. 

```{r}
d %>%
  group_by(flatroof) %>%
  summarise(mean=mean(rainfall))
```

## Extreme values
What are the extreme values? `getOutliers()` computes extreme values against a baseline expectation of a normal distribution

```{r}
outlierPlot(d$rainfall, getOutliers(d$rainfall), title="Rainfall — QQ plot with outliers\nnormal distribution, R2 = 0.9764") 

```

This QQ plot based on a normal distribution doesn't look optimal: two extreme values are classified as outliers, together with the tail end of what looks like a continuous distribution. Can we do better?

Well, precipitation is probably better modelled using a Weibull distribution: after all, 0 is a hard minimum whereas there is more room for variation on the other end (cf. Wilks 1989). Whoah, that's a pretty tight fit — R² = 0.9988!

```{r}
outlierPlot(d$rainfall, getOutliers(d$rainfall,distribution="weibull"), title="Rainfall — QQ plot with outliers\nWeibull distribution, R² = 0.9988") 
```

So what are those extreme values? They have exactly the same rainfall and seem to be very close together: Saipan Island in the NW Pacific.
```{r}
d[c(getOutliers(d$rainfall,distribution="weibull")$iRight),c("Preferred.society.name","region","rainfall","latitude","longitude")]
```

Monthly rainfall for both is recorded as 547.9482, probably taken or interpolated from the same weather station(s). The lat/long places both on Saipan Island. Annual rainfall there is highly variable, but ranges from 1900mm to 2300mm, i.e. 158-192 monthly, according to a 2004 report [1] — three times lower than the 548 monthly from this estimate. It is hard to find sources with historical precipitation data for the focal years of the Ethnographic Atlas (1930-1950), but I consider this enough reason to treat these data points as quirks to be excluded.

[1] http://www.weriguam.org/reports/item/rainfall-climatology-for-saipan-distribution-return-periods-el-nino-tropical-cyclones-and-long-term-variations.html

Though we'll see below that the result don't change whether these two outliers are included or not, we'll work with a copy of the data that does excude them, mostly because otherwise the colour scales of the plots are off.

```{r}
d.no <- d[-getOutliers(d$rainfall,distribution="weibull")$iRight,] %>%
  group_by(flatroof) 
d.no %>% summarise(count=n(),mean=mean(rainfall))

```

So what are the 3 societies with the highest rainfall in the flat roof category?

```{r}
top3 <- d.no %>%
  filter(flatroof == 1) %>%     # filter for flat roofs
  arrange(desc(rainfall)) %>%   # arrange to sort by rainfall
  slice(1:3)                    # slice to take the first 3 rows
top3[,c("Preferred.society.name","region","rainfall","roofdesc","references")]
```

# Visualising


## Plotting
Basic plots
```{r}
myseed=100
set.seed(myseed)
ggplot(d.no, aes(x=flatroof,y=rainfall,fill=rainfall,colour=rainfall)) +
  theme_bw() +
  geom_jitter(alpha=0.7,shape=21,width=0.5,size=2,stroke=1) +
  scale_fill_gradient("rainfall", low="#5edcff", high="#035280") +
  scale_colour_gradient("rainfall", low="#5edcff", high="#035280") +
  stat_summary(fun.y = "median", fun.ymin = "median", fun.ymax= "median", size=0.3,width=0.33, geom = "crossbar")
ggsave(file="out/rainfall-by-rooftype.png", width=3,height=5)

# Highlight highest 3
d.no$top3 <- ifelse(d.no$Preferred.society.name %in% top3$Preferred.society.name, 1, NA)
d.no$top3 <- as.factor(d.no$top3)

set.seed(myseed)
ggplot(d.no, aes(x=flatroof,y=rainfall,fill=top3,colour=rainfall)) +
  theme_bw() +
  geom_jitter(alpha=0.7,shape=21,width=0.5,size=2,stroke=1) +
  scale_colour_gradient("rainfall", low="#5edcff", high="#035280") +
  scale_fill_manual(values=c("red"), guide = FALSE) +
  stat_summary(aes(x=flatroof,y=rainfall),inherit.aes=F,fun.y = "median", fun.ymin = "median", fun.ymax= "median", size=0.3,width=0.33, geom = "crossbar")
 ggsave(file="out/rainfall-by-rooftype-top3.png", width=3,height=5)

```
## Maps
Maps confirm that flat roofs predominantly occur in drier places.

```{r}
ggplot(d.no,aes(longitude,latitude)) + 
  theme_map() +
  borders("world", colour=NA, fill="#e7e8ea") +
  geom_point(data=filter(d.no,flatroof==0),aes(fill=rainfall,colour=flatroof),size=3,shape=21,stroke=1,alpha=0.7) +
  geom_point(data=filter(d.no,flatroof==1),aes(fill=rainfall,colour=flatroof),size=3,shape=22,stroke=1,alpha=0.7) +
  scale_colour_manual(values = c("#ffffff","brown"), guide = FALSE) +
  scale_fill_gradient("rainfall", low="#5edcff", high="#035280")
ggsave(file="out/map-rainfall-by-rooftype.png", width=12,height=8)

# map highest
last_plot() + geom_point(data=top3,aes(longitude,latitude),fill="red",colour="red",size=3,shape=22,stroke=1)
ggsave(file="out/map-rainfall-by-rooftype-highest.png", width=12,height=8)

# zoom: West Africa
last_plot() + theme_bw() # use to target desired lat/long for zooming

zoom1 <- last_plot() + borders(colour="white") + coord_map(xlim = c(-18,7), ylim = c(20,2))

zoom1 + geom_point(data=top3,aes(longitude,latitude),fill="red",colour="red",size=3,shape=22,stroke=1)
ggsave(file="out/map-rainfall-by-rooftype-west-africa.png", width=12,height=8)

# quick look at duplicates due to D-PLACE lat/long binning
d <- d %>% 
  mutate(latlong = paste0("Lo",longitude,"La",latitude),
         dup = duplicated(latlong))
duplatlongs <- subset(d,dup==T)$latlong

# looks like 240 societies have non-unique lat/long values
d %>%
  filter(latlong %in% duplatlongs) %>%
  arrange(latlong) %>%
  group_by(latlong) %>%
  select(Preferred.society.name,latitude,longitude) %>%
  View()

```

# Stats
So the difference in means and in distributions is suggestive. But not all observations are independent. For one thing, some languages are from the same language family and so may have inherited cultural features from a common ancestor. We use generalised linear mixed effects modelling to control for this. We add language family as a random intercept (meaning we allow the value of roof to be 'dependent' to some degree on language family). 

We create two models: one in which we try to predict flatroof by language family alone, another where we add rainfall as a fixed effect. The second one is what we're primarily interested in. Comparing them, the one including rainfall as a fixed effect, in addition to language family, explains the spread of data significantly better (χ2(1) = 51.331, p < 0.00001, log likelihood difference 25.67).


```{r}
model1 <- glmer(flatroof ~ (1|Language.family), data=d.no,family=binomial)
summary(model1)
model2 <- glmer(flatroof ~ rainfall + (1|Language.family), data=d.no,family=binomial)
summary(model2)

comparison <- anova(model1, model2)
comparison[1,4]-comparison[2,4] # log likelihood difference
```
But the maps show that some of the flat roofs appear in 'clumps'. Perhaps horizontal diffusion due to cultural contact can explain part of it. Best way to check this would be to compute distance to nearest flatroof neighbour, but the coordinate data in D-PLACE is rounded which makes that measure uninformative in regions with many societies. So let's use region as a proxy. The reasoning is: if you're in the same region, chances are you have the same kind of roof, whatever it is (i.e. a random intercept). 

We compare model 2 with this new model. The new model, which includes region besides language family, explains the spread of data significantly better (χ2(1) = 56.262, p < 0.00001, log likelihood difference 28.13).

```{r}
model3 <- glmer(flatroof ~ rainfall + (1|region) + (1|Language.family), data=d.no,family=binomial)
summary(model3)

comparison <- anova(model2, model3)
comparison[1,4]-comparison[2,4] # log likelihood difference
```

Do we even need rainfall, or are region and language family sufficient to explain the distribution of flat roofs? To find out, we compare a model with just region & language family as random effects (model 4) with a model with rainfall added as a fixed effect (model 3). The fullest model, with rainfall as main predictor and region and language family as additional factors, explains the data significantly better (χ2(1) = 29.704, p < 0.00001, log likelihood difference 14.85).

```{r}
model4 <- glmer(flatroof ~ (1|region) + (1|Language.family), data=d.no,family=binomial)
summary(model4)

comparison <- anova(model3, model4)
comparison
comparison[1,4]-comparison[2,4] # log likelihood difference
```

Footnote: We excluded some extreme values (which turned out to be in the flatroof=0 category). Would including them have made any difference for the results? Nope: the full model (model 6) explains the data significantly better (χ2(1) = 29.734, p < 0.00001, log likelihood difference 14.87). Excluding the extreme values was mainly useful to make the plots a little prettier.
```{r}
model5 <- glmer(flatroof ~ (1|region) + (1|Language.family), data=d,family=binomial)
model6 <- glmer(flatroof ~ rainfall + (1|region) + (1|Language.family), data=d,family=binomial)
comparison <- anova(model5, model6)
comparison
comparison[1,4]-comparison[2,4] # log likelihood difference
```

# Botero (2014) Nearest-neighbour approach

Here is an approach implemented first in 'The ecology of religious beliefs' by Carlos Botero and colleagues. This approach includes language family as a random effect (as in the model above) and estimates the effect of spatial proximity by using the average response value of N nearest neighbours as a predictor variable. The logic being if something is transmitted horizontally, neighbours have similar trait values and this will predict a particular societies value. 

First, we will look at the NN approach - however, since we are using a global sample, neighbours are not always of the same distance, meaning the definition of horiztonal influence varies (e.g. Rapanui's nearest 5 neighbours will be very different to a PNG society). So, secondly we alter this approach to take the nearest neighbours within a certain distance. This approach will be called closest neighbours. 

Latitude and Longitude variables are used to calculate nearest & closest neighbours. 

## Nearest neighbour

The default value of this approach is to use `NN = 10` which we have used here. The model here is effectively, `flatroof ~ rainfall + NN Score + (1|Language Family)`

```{r, message=FALSE, warning=FALSE}
d$flatroof = d$flatroof %>% 
  as.character(.) %>% 
  as.numeric(.)

mydata = d
response = "flatroof"
Predictors = "rainfall"
VerticalDependency = "Language.family"
Nneighbors = 1000

fit = MuMAC(mydata = d, 
            response = "flatroof", 
            Predictors = "rainfall", 
            VerticalDependency = "Language.family", 
            Nneighbors = 10)
#fit

kable(fit$ModWeights, format = 'html', digits = 3) %>% 
  kable_styling()
kable(fit$MM_average, format = 'html', digits = 3) %>% 
  kable_styling()
```

This approach fits all possible models, and compares them based on AICc values. The first table here shows the delta AICc and AICc weight of each model. The second table shows the beta estimates for each variable in the model and the Relative Variable Importance (RVI). RVI is calculated using the presence of variables in the better fitting models - i.e. if a variable is often in the better fitting models, then it has high RVI. Predictive accuracy is another output from this model, but it hasn't been utilised here so is empty. These outputs are the same for the closest neighbours output. 

Using the default NN value, we see the best model is one only using the Nearest neighbour score. Neighbour score also has an RVI of one. (The intercept has to be in every model). This contradicts the ideas above. 

```{r}
fit = MuMAC(mydata = d, 
            response = "flatroof", 
            Predictors = "rainfall", 
            VerticalDependency = "Language.family", 
            Nneighbors = 20)
#fit

kable(fit$ModWeights, format = 'html', digits = 3) %>% 
  kable_styling()
kable(fit$MM_average, format = 'html', digits = 3) %>% 
  kable_styling()
```

Using double the number of Neighbours, we see the importance of Vertical dependency increases to a much higher level. This is perhaps because NN scores now spans across language families. Rainfall however, remains relatively unimportant. The most important model now is one including NeighbourScore and the Language family random effect. 

## Closest neighbours

Here, I have altered the approach slightly to take the average of neighbours within a certain distance. _This is currently a small hack in the code which mean `Nneighbours` now represents distance rather than number of neighbours_. Currently I am just guesstimating the best distance, so a more methodological approach to choosing distance might be appropriate. I will start with neighbours within 1000kms. 

```{r, message=FALSE, warning=FALSE}
fit = MuMAC_distance(mydata = d, 
            response = "flatroof", 
            Predictors = "rainfall", 
            VerticalDependency = "Language.family", 
            Nneighbors = 1000) 
# Nneighbours now takes a distance value for neighbours to be within

kable(fit$ModWeights, format = 'html', digits = 3) %>% 
  kable_styling()
kable(fit$MM_average, format = 'html', digits = 3) %>% 
  kable_styling()
```

When we run this approach all predictors are of basically equal importance, and therefore are all included in the best model.

If we reduce the NN distance to 500km

```{r, message=FALSE, warning=FALSE}
fit = MuMAC_distance(mydata = d, 
            response = "flatroof", 
            Predictors = "rainfall", 
            VerticalDependency = "Language.family", 
            Nneighbors = 500) 
# Nneighbours now takes a distance value for neighbours to be within

kable(fit$ModWeights, format = 'html', digits = 3) %>% 
  kable_styling()
kable(fit$MM_average, format = 'html', digits = 3) %>% 
  kable_styling()
```

What is not clear is how good this model is to the Nearest neighbour models. This will need further investigating. 


# Conclusions
In this dataset, rainfall is a strong predictor of roof shape, controlling for language family and region. Flat roofs are unlikely to occur in locales with a lot of precipitation. So the hypothesis is supported: rainfall selects against flat roots.

## References

Botero, C. A., Gardner, B., Kirby, K. R., Bulbulia, J., Gavin, M. C., & Gray, R. D. (2014). The ecology of religious beliefs. Proceedings of the National Academy of Sciences, 111(47), 16784–16789. https://doi.org/10.1073/pnas.1408701111
